---
title: "EMIS timeseries"
output:
  html_document:
    df_print: paged
---

# Introduction

EMISR is a package to download RS dataset from EMIS website. 

## Products

There are several products available. We can retrieve the complete list and see which are most useful for the particular purpose.

```{r echo=FALSE}

require(EMISR)

# read available datasets at 4 and 2km resolution

data_emis_4km<-read_emis_wcst_data(resolution="4km",emis_wcst_url="http://emis.jrc.ec.europa.eu/webservices/4km/wcs-t")

data_emis_2km<-read_emis_wcst_data(resolution="2km",emis_wcst_url="http://emis.jrc.ec.europa.eu/webservices/2km/wcs-t")
  
```

So far we are interested in several datasets:

- MODIS-T 4km SST (degree-C)

- MODIS-A 4km CHLA (log10 mg.m^-3)

- ANOMALIES

- MODIS-A 4km PAR

- PAS, OXYRISK (link)[http://mcc.jrc.ec.europa.eu/emis/dev.py?N=simple&O=292&titre_page=Definition%20of%20eutrophication&titre_chap=EUTROPHICATION]


## Details

2Km and 4km resolutions are available.

We can retrieve data products details:

4km

```{r}

SST4KM <- data_emis_4km[data_emis_4km$name=="MODIS-T 4km SST",]
SSTANO4KM <- data_emis_4km[data_emis_4km$name== "MODIS-T 4km SST Anomalies",]
CHLA4KM <- data_emis_4km[data_emis_4km$name=="MODIS-A 4km CHLA",]
CHLAANO4KM <- data_emis_4km[data_emis_4km$name=="MODIS-A 4km CHLA Anomalies",]

PAR4KM <- data_emis_4km[data_emis_4km$name=="MODIS-A 4km PAR",]
OXYRISK4KM <- data_emis_4km[data_emis_4km$name=="MODEL 4km OXYRISK",]
PSA4KM <- data_emis_4km[data_emis_4km$name=="MODEL 4km PSA",]


```

2km

```{r}

CHLA2KM <- data_emis_2km[data_emis_2km$name=="MODIS-A 2km CHLA",]
SST2KM <- data_emis_2km[data_emis_2km$name=="MODIS-T 2km SST",]
PAR2KM <-  data_emis_4km[data_emis_4km$name=="MODIS-A 2km PAR",]

```


### Extent

The function to download datasets require a bounding box. We use both  EZZ and territorial seas, but the former is used. As we pointed out before, we crop this raster stack later using a lower bounding box.  

EZZ 

```{r}
EZZ.mad <- st_read(dsn = '../../../GIS/Marine GIS/output/EEZ_Madeira.shp')
EZZ.mad.box <- st_bbox(EZZ.mad)
#plot(EZZ.mad)
EZZ.mad.box
```

Territorial Seas

```{r echo=FALSE}
require(sf)
ts.mad <- st_read(dsn = '../../../GIS/Marine GIS/output/territorial_seas_12nm_madeira.shp')
ts.mad.box <- st_bbox(ts.mad)
#plot(territorial.seas)
ts.box
```


EZZ bounding box. Uncomment to get the TS bounding box.

```{r}
emisRbbox <- EZZ.mad.box
#emisRbbox <- ts.bbox
```


### EMIS DATA 4km

Get 4km dataset: SST, CHLA and PAR. Other datasets are available but there are errors. 

```{r}


if (!file.exists('output/SST4KM.tif')) {
  
  attach(SST4KM)
  
  SST4KM.md <- getemisdataseries(name = shortname, 
                                 resolution = as.character(resolution), 
                                 startdate = startdate, 
                                 enddate = enddate, 
                                 xmin = emisRbbox["xmin"], xmax = emisRbbox["xmax"], 
                                 ymin = emisRbbox["ymin"], ymax = emisRbbox["ymax"])
  
  writeRaster(SST4KM.md, 'output/SST4KM.tif', overwrite=TRUE)
  
  detach(SST4KM)
  
  }


if (!file.exists('output/SSTANO4KM.tif')) {
  
  attach(SSTANO4KM)
  
  SSTANO4KM.md <- getemisdataseries(name = shortname, 
                                 resolution = as.character(resolution), 
                                 startdate = startdate, 
                                 enddate = enddate, 
                                 xmin = emisRbbox["xmin"], xmax = emisRbbox["xmax"], 
                                 ymin = emisRbbox["ymin"], ymax = emisRbbox["ymax"])
  
  writeRaster(SSTANO4KM.md, 'output/SSTANO4KM.tif', overwrite=TRUE)
  
  detach(SSTANO4KM)
  
}


if (!file.exists('output/CHLA4KM.tif')) {

  attach(CHLA4KM)
  
  CHLA4KM.md <- getemisdataseries(name = shortname, 
                                  resolution = as.character(resolution), 
                                  startdate = startdate, 
                                  enddate = enddate, 
                                  xmin = emisRbbox["xmin"], xmax = emisRbbox["xmax"], 
                                  ymin = emisRbbox["ymin"], ymax = emisRbbox["ymax"])
  
  writeRaster(CHLA4KM.md, 'output/CHLA4KM.tif', overwrite=TRUE)
  
  detach(CHLA4KM)
}


if (!file.exists('output/CHLAANO4KM.tif')) {

  attach(CHLAANO4KM)
  
  CHLAANO4KM.md <- getemisdataseries(name = shortname, 
                                  resolution = as.character(resolution), 
                                  startdate = startdate, 
                                  enddate = enddate, 
                                  xmin = emisRbbox["xmin"], xmax = emisRbbox["xmax"], 
                                  ymin = emisRbbox["ymin"], ymax = emisRbbox["ymax"])
  
  writeRaster(CHLAANO4KM.md, 'output/CHLAANO4KM.tif', overwrite=TRUE)
  
  detach(CHLAANO4KM)
}

if (!file.exists('output/PAR4KM.tif')) {


  attach(PAR4KM)
  
  PAR4KM.md <- getemisdataseries(name = shortname, 
                                 resolution = as.character(resolution), 
                                 startdate = startdate, 
                                 enddate = enddate, 
                                 xmin = emisRbbox["xmin"], xmax = emisRbbox["xmax"], 
                                 ymin = emisRbbox["ymin"], ymax = emisRbbox["ymax"])
  
  writeRaster(PAR4KM.md, 'output/PAR4KM.tif',  overwrite=TRUE)
  
  detach(PAR4KM)
}

if (!file.exists('output/PSA4KM.tif')) {
  
  attach(PSA4KM)
  
  PSA4KM.md <- getemisdataseries(name = shortname, 
                                 resolution = as.character(resolution), 
                                 startdate = startdate, 
                                 enddate = enddate, 
                                 xmin = emisRbbox["xmin"], xmax = emisRbbox["xmax"], 
                                 ymin = emisRbbox["ymin"], ymax = emisRbbox["ymax"])
  
  writeRaster(PSA4KM.md, 'output/PSA4KM.tif')
  
  detach(PSA4KM)
  
  }

if (!file.exists('output/OXYRISK4KM.tif')) {
  
  attach(OXYRISK4KM)
  
  OXYRISK4KM.md <- getemisdataseries(name = shortname, 
                                 resolution = as.character(resolution), 
                                 startdate = startdate, 
                                 enddate = enddate, 
                                 xmin = emisRbbox["xmin"], xmax = emisRbbox["xmax"], 
                                 ymin = emisRbbox["ymin"], ymax = emisRbbox["ymax"])
  
  writeRaster(OXYRISK4KM.md, 'output/OXYRISK4KM.tif')
  
  detach(OXYRISK4KM)
  
  }

```


### EMIS DATA 2km

Equal to 2km datasets



```{r}
#emisRbbox <- ts.bbox
emisRbbox <- EZZ.mad.box

```


```{r}

if (!file.exists('output/SST2KM.tif')) {
  
  attach(SST2KM)
  
  SST2KM.md <- getemisdataseries(name = shortname, 
                                 resolution = as.character(resolution), 
                                 startdate = startdate, 
                                 enddate = enddate, 
                                 xmin = emisRbbox["xmin"], xmax = emisRbbox["xmax"], 
                                 ymin = emisRbbox["ymin"], ymax = emisRbbox["ymax"])
  
  writeRaster(SST2KM.md, 'output/SST2KM.tif')
  
  detach(SST2KM)
  
  }


if (!file.exists('output/CHLA2KM.tif')) {

  attach(CHLA2KM)
  
  CHLA2KM.md <- getemisdataseries(name = shortname, 
                                  resolution = as.character(resolution), 
                                  startdate = startdate, 
                                  enddate = enddate, 
                                  xmin = emisRbbox["xmin"], xmax = emisRbbox["xmax"], 
                                  ymin = emisRbbox["ymin"], ymax = emisRbbox["ymax"])
  
  writeRaster(CHLA2KM.md, 'output/CHLA2KM.tif')
  
  detach(CHLA2KM)
}

if (!file.exists('output/PAR2KM.tif')) {


  attach(PAR2KM)
  
  PAR2KM.md <- getemisdataseries(name = shortname, 
                                 resolution = as.character(resolution), 
                                 startdate = startdate, 
                                 enddate = enddate, 
                                 xmin = emisRbbox["xmin"], xmax = emisRbbox["xmax"], 
                                 ymin = emisRbbox["ymin"], ymax = emisRbbox["ymax"])
  
  writeRaster(PAR2KM.md, 'output/PAR2KM.tif')
  
  detach(PAR2KM)
}

```


### Crop

A new bounding box is produced to crop the previously downloaded datasets. 

This bounding box is done by buffering the coast line using a distance of 5Km


```{r}

require(sf)

cl.pol.mad <- st_read('output/cl_pol_mad.shp')

cl.pol.mad <- st_cast(foo, 'POLYGON')

i <- which(st_area(cl.pol.mad)==max(st_area(cl.pol.mad))) # identify main island by area

mad.cl <- cl.pol.mad$geometry[i] # get the main polygon

mad.cl.sp <- as(mad.cl, "Spatial")

mad.bb <-st_bbox(mad.cl)

mad.bb.buff <- st_buffer(st_transform(st_as_sfc(mad.bb), 2942), dist = 5000)

mad.bb.sp <- as((st_as_sfc(st_bbox(st_transform(mad.bb.buff, 4326)))), "Spatial") # a bounding box 
mad.bb.sf <- st_as_sf(mad.bb.sp)

st_write(mad.bb.sf, "output/mad_bb.shp")
```

