---
title: "EMIS SST"
output:
  html_document:
    df_print: paged
---

# SST (4KM)

## Extent

Get the previously computed bounding box to crop raster stack

```{r}
mad.bb.sf <- st_read("output/mad_bb.shp")
mad.bb.sp <- as(mad.bb.sf, "Spatial")
```


## Crop

Crop, mask, save or read multitemporal SST4KM cropped raster stack

```{r}

if(!file.exists('output/sst4km_crop.tif')) {
  SST4KM <- stack('output/SST4KM.tif')
  SST4KM.crop <- raster::crop(x = SST4KM, y = mad.bb.sp )
  SST4KM.crop <- mask(SST4KM.crop, mask = mad.cl.sp, inverse=T)
  writeRaster(SST4KM.crop, 'output/sst4km_crop.tif', overwrite=T)
}  else {
  SST4KM.crop <- stack('output/sst4km_crop.tif')

}

dim(SST4KM.crop)
print(SST4KM.crop)

```

```{r}
plot(SST4KM.crop[[208]])
```

##  Timeseries


### Build temporal framework
Check time series 

http://mcc.jrc.ec.europa.eu/emis/dev.py?N=50&O=306&titre_page=4km%20Marine%20Data&titre_chap=Data%20discovery

EMIS_T_SST_MM_YYYY

2000 Include January

2017 Include from June

```{r}

buildStack <- function(raster, n) {
  r <- raster[[1]] # get first layer
  r[] <- NA # set NA values
  r <- stack(replicate(n, r)) # replicate based on missing temporal dataset
}

if(!file.exists('r1SST.rds')) {

seqt.2000 <- seq.Date(from = as.Date("2000-01-31"), 
                      to = as.Date("2000-01-31"), 
                      by = "month", lenght.out=12) # seq for missing months

seqt.2017 <- seq.Date(from = as.Date("2017-06-01"), 
                      to = as.Date("2017-12-01"), 
                      by = "month", lenght.out=12) # equal

seqt.2000_2017 <- seq.Date(from = as.Date("2000-01-01"), 
                           to = as.Date("2017-12-31"), 
                           by = "month", lenght.out=12) # whole time serie

r1 <- buildStack(SST4KM.crop, 1)
r2 <- buildStack(SST4KM.crop, 7)


r1SST <- stack(r1, SST4KM.crop) # fill with 2002 
r1SST <- stack(r1SST, r2) # fill above with 2017

r1SST <- setZ(r1SST, seqt.2000_2017) # set z
saveRDS(r1SST, "r1SST.rds")

} else {
  
  r1SST <-readRDS('r1SST.rds')
}
```


### Summarize 


```{r}
require(raster)
require(zoo) 
require(lubridate) 
```


#### Year

```{r}

indices <-  format(as.Date(getZ(r1SST), format = "X%Y.%m.%d"), format = "%y") 
indices <- as.numeric(indices)
```


##### Mean

```{r}



SST4KM.crop.year.mean <-  stackApply(r1SST, indices , fun = mean)
plot(SST4KM.crop.year.mean)

```


```{r}
if(!file.exists('output/sst4km_crop_year_mean.tif'))  writeRaster(SST4KM.crop.year.mean, 'output/sst4km_crop_year_mean.tif', overwrite=T)

```


##### Max

```{r}

SST4KM.crop.year.max <-  stackApply(r1SST, indices , fun = max)
plot(SST4KM.crop.year.max)

```

```{r}
if(!file.exists('output/sst4km_crop_year_max.tif'))  writeRaster(SST4KM.crop.year.max, 'output/sst4km_crop_year_max.tif', overwrite=T)

```


##### Min

```{r}

SST4KM.crop.year.min <-  stackApply(r1SST, indices , fun = min)
plot(SST4KM.crop.year.min)

```


```{r}
if(!file.exists('output/sst4km_crop_year_min.tif'))  writeRaster(SST4KM.crop.year.min, 'output/sst4km_crop_year_min.tif', overwrite=T)

```



#### Month

```{r}

indices <-  format(as.Date(getZ(r1SST), format = "X%Y.%m.%d"), format = "%m") 
indices <- as.numeric(indices)

```


##### Mean
```{r}

SST4KM.crop.month.mean <-  stackApply(r1SST, indices , fun = mean)
plot(SST4KM.crop.month.mean)

```

```{r}
if(!file.exists('output/sst4km_crop_month_mean.tif'))  writeRaster(SST4KM.crop.month.mean, 'output/sst4km_crop_month_mean.tif', overwrite=T)

```


```{r}
SST4KM.crop.month.mean 
```


##### Max

```{r}

SST4KM.crop.month.max <-  stackApply(r1SST, indices , fun = max)
plot(SST4KM.crop.month.max)
```


```{r}
if(!file.exists('output/sst4km_crop_month_max.tif'))  writeRaster(SST4KM.crop.month.max, 'output/sst4km_crop_month_max.tif', overwrite=T)

```


##### Min

```{r}

SST4KM.crop.month.min <-  stackApply(r1SST, indices , fun = min)
plot(SST4KM.crop.month.min)
```


```{r}
if(!file.exists('output/sst4km_crop_month_min.tif'))  writeRaster(SST4KM.crop.month.min, 'output/sst4km_crop_month_min.tif', overwrite=T)

```


## Results

Once raster timeseries has been reduced to monthly and yearly datasets it is time to explore the timeseries. 


### Month

#### Mean

##### Summarize by raster


Bind values in a whole dataframe

```{r}
require(reshape2)

SST4KM.crop.month.mean.df <- as.data.frame(SST4KM.crop.month.mean)


names(SST4KM.crop.month.mean.df) <- c(1:12)

SST4KM.crop.month.mean.df <- melt(SST4KM.crop.month.mean.df) # melt values and months


monthAsFactor <- factor(SST4KM.crop.month.mean.df$variable, labels = c("Jan", "Feb", "Mar", "Apr", "May","Jun", "Jul", "Aug", "Sep","Oct", "Nov", "Dec"), ordered = FALSE)


SST4KM.crop.month.mean.df$month <-  monthAsFactor


```


```{r}

if (!file.exists('output/sst4km_crop_month_mean.txt')){
  write.table(SST4KM.crop.month.mean.df, 
              'output/sst4km_crop_month_mean.txt', row.names = FALSE, col.names = TRUE, sep = ",")
} 
```

```{r}
ggplot(SST4KM.crop.month.mean.df) + 
  geom_boxplot(aes(x = month, y = value)) +
  theme_bw()
```


##### SST peak (SSTP) seasonal value (mean based)

SST peak (SSTP) seasonal value as the average of SST mean values in August and September.


```{r}
SST4KM.crop.month.mean.aug.sep <- overlay(SST4KM.crop.month.mean$SST4KM_Aug, SST4KM.crop.month.mean$SST4KM_Sep, fun=mean)

writeRaster(SST4KM.crop.month.mean.aug.sep, 'output/SSTP_Mean_Seasonal_0809m_0017Y.tif')

```

```{r}
plot(SST4KM.crop.month.mean.aug.sep)
```



```{r}
require(dplyr)

SST4KM.crop.month.mean.aug.sep.df <- as.data.frame(SST4KM.crop.month.mean.aug.sep)
max(SST4KM.crop.month.mean.aug.sep.df, na.rm=T)

```


##### Summarize by scalar (single value)


```{r}
SST4KM.crop.month.mean.n <- cellStats(SST4KM.crop.month, stat='mean', na.rm=TRUE)
SST4KM.crop.month.median.n <- cellStats(SST4KM.crop.month, stat='median', na.rm=TRUE)
SST4KM.crop.month.max.n <- cellStats(SST4KM.crop.month, stat='max', na.rm=TRUE)
SST4KM.crop.month.min.n <- cellStats(SST4KM.crop.month, stat='min', na.rm=TRUE)
SST4KM.crop.month.sum.n <- cellStats(SST4KM.crop.month, stat='sum', na.rm=TRUE)
SST4KM.crop.month.sd.n <- cellStats(SST4KM.crop.month, stat='sd', na.rm=TRUE)
SST4KM.crop.month.skew.n <- cellStats(SST4KM.crop.month, stat='skew', na.rm=TRUE)
SST4KM.crop.month.rms.n  <- cellStats(SST4KM.crop.month, stat='rms', na.rm=TRUE)

SST4KM.crop.month.stat.df <-  data.frame(mean=SST4KM.crop.month.mean.n, median=SST4KM.crop.month.median.n, max = SST4KM.crop.month.max.n, min = SST4KM.crop.month.min.n, sum = SST4KM.crop.month.sum.n, sd = SST4KM.crop.month.sd.n, skew = SST4KM.crop.month.skew.n, rms = SST4KM.crop.month.rms.n)
SST4KM.crop.month.skew
rownames(SST4KM.crop.month.stat.df) <- c("Jan", "Feb", "Mar", "Apr", "May","Jun", "Jul", "Aug", "Sep","Oct", "Nov", "Dec")
SST4KM.crop.month.stat.df <- data.frame(month = row.names(SST4KM.crop.month.stat.df), SST4KM.crop.month.stat.df)

```


```{r}

write.table(SST4KM.crop.month.stat.df, "output/sst4km_crop_month_stat.txt", sep=',', row.names = FALSE)

```

Plot result

```{r}
plot(SST4KM.crop.month.mean.n)
lines(SST4KM.crop.month.mean.n)
```

Order by min values

```{r}
sort(SST4KM.crop.month.mean.n)
```

February, March, April, January are the months with minimun values 


Typical minimun:


```{r}
SST4KM.tm <-mean(sort(SST4KM.crop.month.mean.n)[1:4])
```

Range:

```{r}
SST4KM.tm.rg <-diff(range(sort(SST4KM.crop.month.mean.n)[1:4]))
```

Nº times min was below min avg:

```{r}

SST4KM.tm.n <- (sum(SST4KM.crop.month.mean.n < SST4KM.tm))

```

```{r}
SST4KM.crop.month.mean[which(SST4KM.crop.month.mean.n < SST4KM.tm)]
```

Sum temp of anomaly: 

```{r}
## CHLA4KM.tm.sum <- sum(sort(CHLA4KM.crop.month.mean)[1:3]) 

SST4KM.tm.sum <- sum(SST4KM.crop.month.mean.n[which(SST4KM.crop.month.mean.n < SST4KM.tm)]) # just one value

```

Nº times min was avobe min avg


Degree heating (day / week / month)

#### Max

##### Summarize by raster

Bind values in a whole dataframe

```{r}
require(reshape2)

SST4KM.crop.month.max.df <- as.data.frame(SST4KM.crop.month.max)

names(SST4KM.crop.month.max.df) <- c(1:12)

SST4KM.crop.month.max.df <- melt(SST4KM.crop.month.max.df) # melt values and months

monthAsFactor <- factor(SST4KM.crop.month.max.df$variable, labels = c("Jan", "Feb", "Mar", "Apr", "May","Jun", "Jul", "Aug", "Sep","Oct", "Nov", "Dec"), ordered = FALSE)

SST4KM.crop.month.max.df$month <-  monthAsFactor

```


```{r}

if (!file.exists('output/sst4km_crop_month_max.txt')){
  write.table(SST4KM.crop.month.max.df, 
              'output/sst4km_crop_month_max.txt', row.names = FALSE, col.names = TRUE, sep = ",")
} 
```

```{r}
ggplot(SST4KM.crop.month.max.df) + 
  geom_boxplot(aes(x = month, y = value)) +
  theme_bw()
```



##### SST peak (SSTP) seasonal value (max based)

SST peak (SSTP) seasonal value as the average of SST max values in August and September.


```{r}
SST4KM.crop.month.max.aug.sep <- overlay(SST4KM.crop.month.max$SST4KM_Aug, SST4KM.crop.month.max$SST4KM_Sep, fun=mean)
plot(SST4KM.crop.month.max.aug.sep)
```


```{r}
writeRaster(SST4KM.crop.month.max.aug.sep, 'output/SSTP_Max_Seasonal_0809m_0017Y.tif')

```

##### Summarize by scalar (single value)


##### Min

##### Summarize by raster

##### Summarize by scalar (single value)



### Final product

Based on boxplots, months are selected 

### Max


```{r}
SST4KM.crop.month.max.mean <- mean(SST4KM.crop.month.max)
plot(SST4KM.crop.month.max.mean)
writeRaster(SST4KM.crop.month.max.max)
```



# SST (2KM)